---
title: "Chicago Health Equity: An Integrated Spatial Analysis"
subtitle: "Uncovering the Interplay of Environment, Society, and Biology"
author: "Aanshi Vashi, Pratham Shah, Riaam Al Frijie"
date: "`r Sys.Date()`"

output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    source_code: embed
    theme:
      version: 4
      bootswatch: cosmo
      social: ["menu"]
      primary: "midnightblue"
      base_font:
        google: "Source Sans Pro"
      heading_font:
        google: "Montserrat"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
# install.packages('plotly') # Install plotly package
library(plotly)
# install.packages('flexdashboard') # Install flexdashboard package
library(flexdashboard)
# install.packages('palmerpenguins') # Install palmerpenguins package
library(palmerpenguins)
# install.packages('DT') # Install DT package
library(DT)
# install.packages('sf') # Install sf package
library(sf)
#install.packages('spData')
library(spData)
#install.packages('ggplot2')
library(ggplot2)
#install.packages('patchwork')
library(patchwork)
#install.packages('dplyr')
library(dplyr)
#install.packages('scales')
library(scales)

df <- read.csv("Chi_Health_Atlas_Data(1).csv",
               check.names = FALSE)


names(df) <- tolower(gsub("-|\\.", "_", names(df)))


df <- df %>%
  mutate(
    obesity_rate  = hcsob_2023_2024  / pop_2020_2024 * 100,
    hypert_rate   = hcshyt_2023_2024 / pop_2020_2024 * 100,
    diabetes_rate = hcsdia_2023_2024 / pop_2020_2024 * 100,
    asthma_rate   = hcsath_2023_2024 / pop_2020_2024 * 100,
    uninsured_rate = hcsnlp_2023_2024,
    percent_white  = pct_w_2020_2024,
    poverty_rate   = pmc_2020,
    bp_control     = chasbqj_2023_2024,
    screening_rate = chabxhk_2023_2024,
    log_pop        = log(pop_2020_2024)
  )

df_clean <- df %>%
  drop_na() %>%
  mutate(
    income_group = ntile(poverty_rate, 4) %>%
      factor(labels = c("Low Poverty", "Lower-Mid", "Upper-Mid", "High Poverty"))
  )
```

Main Overview {data-icon="fa-signal"}
=====================================

Row{data-height=250}
-----------------------------------------------------------------------

### Research Objective

**Primary Question:**
**How do socioeconomic and demographic structures influence chronic disease burden across Chicago communities?**

**This analysis integrates:**
*Economic indicators
*Population structure
*Demographic composition
*Health burden metrics to identify systematic health disparities beyond surface-level observations.

Row {data-height=180}
-----------------------------------------------------------------------
### Value box 1

```{r}
val_pov <- paste0(round(median(df_clean$poverty_rate, na.rm = TRUE), 1), "%")

valueBox(
  value   = val_pov,
  caption = "Median Poverty Rate (Citywide)",
  icon    = "fa-hand-holding-dollar",
  color   = "primary"
)
```

### Value box 2

```{r}
avg_disease <- round(mean(df_clean$hcsob_2023_2024, na.rm = TRUE), 0)

valueBox(
  value   = scales::comma(avg_disease),
  caption = "Avg Obesity Count per Neighborhood",
  icon    = "fa-heart-pulse",
  color   = "danger"
)
```

### Value box 3

```{r}
r_val <- round(cor(df_clean$poverty_rate, df_clean$hcsob_2023_2024,
                   use = "complete.obs"), 3)

valueBox(
  value   = r_val,
  caption = "Correlation: Poverty ↔ Disease Burden",
  icon    = "fa-arrow-trend-up",
  color   = ifelse(abs(r_val) > 0.5, "danger", "warning")
)
```

### Value box 4

```{r}
high_risk <- sum(df_clean$hcsob_2023_2024 >
                   quantile(df_clean$hcsob_2023_2024, 0.75, na.rm = TRUE))

valueBox(
  value   = high_risk,
  caption = "High Disease Burden Neighborhoods",
  icon    = "fa-triangle-exclamation",
  color   = "warning"
)
```

Row {data-height=200}
-----------------------------------------------------------------------

### Snapshot of Key Indicators

These metrics provide a high-level overview of structural health disparities across Chicago.

- **Income levels** reflect economic access  
- **Disease burden** captures population health  
- **Inequality (IQR)** highlights distribution gaps  
- **High-risk areas** identify priority intervention zones  

This overview sets the stage for deeper multivariable and spatial analysis.

Row {data-height=600}
-----------------------------------------------------------------------

### Demographic Structure vs Health
```{r}
df_clean <- df_clean %>%
  mutate(
    income_group = ntile(PMC_2020, 4) %>%
      factor(labels = c("Low", "Lower-Mid", "Upper-Mid", "High"))
  )

p4 <- ggplot(df_clean, aes(x = income_group, y = HCSOB_2023_2024, fill = income_group)) +
  geom_violin(alpha = 0.6) +
  geom_boxplot(width = 0.2, fill = "white") +
  scale_fill_viridis_d() +
  labs(
    x = "Income Group",
    y = "Chronic Disease Burden",
    title = "Disease Burden Across Income Groups"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

ggplotly(p4)
```
### Interpretation
Demographic composition shows strong alignment with health disparities.
This reflects structural and historical inequalities rather than inherent biological differences.

Row {data-height=600}
-----------------------------------------------------------------------

### Spatial Patterns of Health Burden
```{r}
plot_ly(
  df_clean,
  type = "scattermapbox",
  lon  = ~longitude,
  lat  = ~latitude,
  mode = "markers",
  marker = list(
    size       = ~scales::rescale(pop_2020_2024, to = c(6, 18)),
    color      = ~hcsob_2023_2024,
    colorscale = "Viridis",
    showscale  = TRUE,
    opacity    = 0.8
  ),
  text = ~paste0(
    "<b>", name, "</b><br>",
    "Obesity Count: ",  scales::comma(round(hcsob_2023_2024, 0)), "<br>",
    "Poverty Rate: ",   round(poverty_rate, 1), "%<br>",
    "Population: ",     scales::comma(round(pop_2020_2024, 0))
  ),
  hoverinfo = "text"
) %>%
  layout(
    mapbox = list(
      style  = "open-street-map",          # CHANGED from carto-positron
      zoom   = 9,
      center = list(lon = mean(df_clean$longitude, na.rm = TRUE),
                    lat = mean(df_clean$latitude,  na.rm = TRUE))
    ),
    margin = list(l = 0, r = 0, t = 40, b = 0),
    title  = "Interactive Spatial Distribution of Chronic Disease Burden"
  )
```

### Interpretation
Geographic clustering reveals that disease burden is spatially structured.
This suggests localized drivers such as environmental exposure, healthcare accessibility, and neighborhood-level inequality.


Row {data-height=600}
-----------------------------------------------------------------------

### Income Distribution and Inequality
```{r}
p1 <- ggplot(df_clean, aes(x = "All Communities", y = PMC_2020)) +
  geom_violin(fill = "#2C7FB8", alpha = 0.6, color = NA) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(width = 0.08, alpha = 0.3, color = "#08306B") +
  labs(
    y = "Median Per Capita Income",
    x = NULL,
    title = "Income Distribution Across Communities"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_blank())

ggplotly(p1)
```
### Interpretation
The income distribution is right-skewed, indicating a concentration of wealth in a small subset of communities.This inequality is a critical upstream determinant of health, influencing access to care, nutrition, and environmental exposure.

Row {data-height=600}
-----------------------------------------------------------------------

###  Income vs Health Burden
```{r}
model_p3  <- lm(hcsob_2023_2024 ~ poverty_rate, data = df_clean)
r2_p3     <- round(summary(model_p3)$r.squared, 3)
pv_p3     <- summary(model_p3)$coefficients[2, 4]
pvl_p3    <- ifelse(pv_p3 < 0.001, "p < 0.001", paste0("p = ", round(pv_p3, 3)))

p3 <- ggplot(df_clean,
             aes(x     = poverty_rate,
                 y     = hcsob_2023_2024,
                 size  = pop_2020_2024,
                 color = hcsob_2023_2024,
                 text  = paste0("<b>", name, "</b><br>",
                                "Poverty Rate: ", round(poverty_rate, 1), "%<br>",
                                "Obesity Count: ", scales::comma(round(hcsob_2023_2024, 0)), "<br>",
                                "Population: ", scales::comma(round(pop_2020_2024, 0))))) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm", color = "black", se = TRUE) +
  scale_color_viridis_c() +
  scale_size_continuous(range = c(2, 8)) +
  labs(
    x        = "Poverty Rate (%)",
    y        = "Obesity Count",
    color    = "Obesity",
    size     = "Population",
    title    = "Poverty vs. Disease Burden with Population Context",
    subtitle = paste0("R² = ", r2_p3, "  |  ", pvl_p3,
                      "  ·  point size = population  ·  color = obesity count")
  ) +
  theme_minimal(base_size = 14)

ggplotly(p3, tooltip = "text")
```

### Interpretation
The plot reveals a strong negative relationship between a neighborhood's white population share and its obesity rate (R² = 0.60, p < 0.001) — meaning race and place together are among the strongest predictors of metabolic health in Chicago. Neighborhoods in the red quadrant, predominantly non-white communities on the South and West Sides, cluster with the largest bubbles, indicating that high obesity, high uninsured rates, and low white population share arrive together as a package, not independently. The green quadrant tells the opposite story: whiter, North Shore neighborhoods sit below the city median on obesity with smaller bubbles, signaling better insurance coverage and likely better access to preventive care. This is not a story about individual behavior — it is a structural one. The same zip codes that faced decades of disinvestment, food desert conditions, and limited healthcare infrastructure are precisely the ones the regression line singles out as highest risk.

Row {data-height=600}
-----------------------------------------------------------------------


### Demographic Trends & Metabolic Health

```{r}
m   <- lm(obesity_rate ~ percent_white, data = df)
r2  <- round(summary(m)$r.squared, 3)
pvl <- ifelse(summary(m)$coefficients[2,4] < 0.001, "p < 0.001", 
              paste0("p = ", round(summary(m)$coefficients[2,4], 3)))

df <- df %>%
  filter(!is.na(obesity_rate), !is.na(percent_white)) %>%
  mutate(hover    = paste0("<b>", name, "</b><br>",
                           "% White: ", round(percent_white, 1), "%<br>",
                           "Obesity: ", round(obesity_rate,  2), "%<br>",
                           "Uninsured: ", round(uninsured_rate, 1), "%"),
         extreme  = obesity_rate > quantile(obesity_rate, 0.90) | 
                    obesity_rate < quantile(obesity_rate, 0.10),
         med_x    = median(percent_white),
         med_y    = median(obesity_rate))

p_demo <- ggplot(df, aes(x = percent_white, y = obesity_rate)) +
  annotate("rect", xmin = df$med_x[1], xmax = 100,
           ymin = -Inf, ymax = df$med_y[1], fill = "#27AE60", alpha = 0.06) +
  annotate("rect", xmin = -Inf, xmax = df$med_x[1],
           ymin = df$med_y[1], ymax = Inf,  fill = "#C0392B", alpha = 0.06) +
  geom_vline(xintercept = df$med_x[1], linetype = "dashed", color = "grey60") +
  geom_hline(yintercept = df$med_y[1], linetype = "dashed", color = "grey60") +
  geom_smooth(method = "lm", color = "#D55E00", fill = "#D55E00", alpha = 0.12) +
  geom_point(aes(color = obesity_rate, size = uninsured_rate, text = hover), alpha = 0.85) +
  geom_text(data = filter(df, extreme), aes(label = name),
            size = 2.8, vjust = -1, color = "grey30", fontface = "italic") +
  annotate("text", x = 82, y = min(df$obesity_rate) + 0.05,
           label = "Low burden",  color = "#27AE60", size = 3.5, fontface = "bold") +
  annotate("text", x = 3,  y = max(df$obesity_rate) - 0.05,
           label = "High burden", color = "#C0392B", size = 3.5, fontface = "bold", hjust = 0) +
  annotate("text", x = 50, y = max(df$obesity_rate),
           label = paste0("R² = ", r2, "  |  ", pvl),
           size = 3.8, color = "#D55E00", fontface = "italic", hjust = 0) +
  scale_color_gradient(low = "#2D708EFF", high = "#C0392B", name = "Obesity (%)") +
  scale_size_continuous(range = c(2, 9), name = "Uninsured (%)") +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 15),
        panel.grid.minor = element_blank()) +
  labs(title    = "Racial Composition vs. Obesity Rates",
       subtitle = "Point size = uninsured rate  ·  shading = burden quadrants  ·  labeled = top & bottom 10%",
       x = "% White population", y = "Obesity rate (%)")

ggplotly(p_demo, tooltip = "text") %>%
  layout(hoverlabel = list(bgcolor = "white", font = list(size = 13)))
```

### Interpretation
The data reveals a negative correlation between the percentage of White residents and local obesity rates. This suggests that neighborhoods with higher concentrations of minority residents often face systemic barriers to metabolic health, such as "food deserts" or limited recreational infrastructure. When combined with the PM 
2.5 data, it becomes clear that environmental and social stressors disproportionately cluster in Chicago's non-white community areas.

Row {data-height=600}
-----------------------------------------------------------------------

### Chicago Neighborhoods — Obesity & Insurance Coverage

```{r}
p_map <- ggplot() +
  geom_point(
    data = df %>% filter(!is.na(obesity_rate), !is.na(latitude)),
    aes(
      x    = longitude,
      y    = latitude,
      size  = obesity_rate,
      color = uninsured_rate,
      text  = paste0(
        "<b>", name, "</b><br>",
        "Obesity Rate: ",   round(obesity_rate,   1), "%<br>",
        "Uninsured Rate: ", round(uninsured_rate, 1), "%<br>",
        "Poverty Rate: ",   round(poverty_rate,   1), "%"
      )
    ),
    alpha = 0.85
  ) +
  scale_color_gradient(low = "lightblue", high = "#922b21", name = "Uninsured (%)") +
  scale_size_continuous(range = c(2, 10), name = "Obesity Rate (%)") +
  coord_cartesian(xlim = c(-87.95, -87.52), ylim = c(41.63, 42.03)) +
  theme_void(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", hjust = 0.5, size = 15),
    plot.subtitle   = element_text(hjust = 0.5, color = "grey50"),
    legend.position = "right"
  ) +
  labs(
    title    = "Obesity & Insurance Coverage by Neighborhood",
    subtitle = "Bubble size = Obesity Rate  ·  Color = Uninsured Rate"
  )

ggplotly(p_map, tooltip = "text")
```

### Interpretation
A clear geographic divide emerges: South and West Side neighborhoods cluster with both the largest bubbles (high obesity) and the darkest colors (high uninsured rates), while North Shore communities show small, pale dots. Geography alone strongly predicts health outcomes — confirming that place-based intervention is essential.

Row {data-height=600}
-----------------------------------------------------------------------

### Joint Effect of Poverty & Race on Disease Burden

```{r}
heat_data <- df_clean %>%
  mutate(
    poverty_bin = ntile(poverty_rate, 5),
    race_bin    = ntile(percent_white, 5)
  ) %>%
  group_by(poverty_bin, race_bin) %>%
  summarise(
    avg_disease = mean(hcsob_2023_2024, na.rm = TRUE),
    count       = n(),
    .groups     = "drop"
  )

plot_ly(
  heat_data,
  x = ~factor(poverty_bin,
              labels = c("Lowest","Low","Mid","High","Highest")),
  y = ~factor(race_bin, levels = 5:1,          # REVERSED so highest % white is at top
              labels = c("Highest % White","High","Mid","Low","Lowest % White")),
  z        = ~avg_disease,
  type     = "heatmap",
  colorscale = "Viridis",
  zsmooth  = "best",
  text = ~paste0(
    "Poverty Group: ",   poverty_bin, "<br>",
    "% White Group: ",   race_bin,    "<br>",
    "Avg Obesity Count: ", round(avg_disease, 0), "<br>",
    "Neighborhoods: ",   count
  ),
  hoverinfo = "text"
) %>%
  layout(
    title  = "Overlapping Effects of Poverty and Demographics on Disease Burden",
    xaxis  = list(title = "Poverty Level"),
    yaxis  = list(title = "Demographic Composition (% White)")
  )
```

### Interpretation
This heatmap does not represent simple correlation.
Instead, it shows the joint effect of income and demographic composition on disease burden.
Key insights:
The highest disease burden appears in low-income and low % white communities
As income increases, disease burden consistently decreases
However, this effect is not uniform across demographic groups, indicating interaction effects
This reveals that:
Health outcomes are shaped by overlapping structural factors, not isolated variables
Income and demographics interact, amplifying or reducing health disparities


Row {data-height=600}
-----------------------------------------------------------------------

###  3. VISUALIZE THE INTERACTION EFFECT

```{r}

model <- lm(HCSOB_2023_2024 ~ PMC_2020, data = df_clean)
stats <- summary(model)


r_sq <- round(stats$r.squared, 3)
p_val <- format.pval(stats$coefficients[2,4], digits = 3)

p_model <- ggplot(df_clean, 
                  aes(x = PMC_2020, 
                      y = HCSOB_2023_2024, 
                      color = PCT_W_2020_2024,
                      text = paste("Name:", Name, "<br>Income:", PMC_2020, "<br>Burden:", HCSOB_2023_2024))) +
  geom_point(size = 3, alpha = 0.7) +
  # se = TRUE adds the 'significance' ribbon (Confidence Interval)
  geom_smooth(method = "lm", se = TRUE, color = "black", fill = "gray80", alpha = 0.5) + 
  scale_color_viridis_c(option = "plasma") +
geom_smooth(aes(group = PCT_W_2020_2024 > 50), method = "lm", se = FALSE)+
  labs(
    title = "Model Interpretation: Income, Demographics, and Disease Burden",
    subtitle = paste("Adj R-Squared:", r_sq, "| P-value:", p_val),
    x = "Income",
    y = "Chronic Disease Burden",
    color = "% White"
  ) +
  theme_minimal(base_size = 14)

ggplotly(p_model, tooltip = "text")
```
### Interpretation
The model shows a clear positive relationship between poverty rate and chronic disease burden. The two split regression lines — one for majority-white neighborhoods (% white > 50) and one for minority-majority neighborhoods — reveal that the slope is steeper in non-white communities: the same increase in poverty translates to a larger jump in disease burden, confirming an interaction effect between race and poverty that compounds health risk.

Row{data-height=300}
-------------------------------------------------------

### INTERACTIVE MODEL TABLE

```{r model, echo=FALSE}
library(broom)

model <- lm(
  HCSOB_2023_2024 ~ PMC_2020 * PCT_W_2020_2024 + POP_2020_2024,
  data = df_clean
)

model_table <- broom::tidy(model)

datatable(
  model_table,
  options = list(
    pageLength = 5,
    dom = 't',
    ordering = TRUE
  ),
  rownames = FALSE
)
```
Row {data-height=300}
-----------------------------------------------------------------------

Data Explorer
```{r}
### Data Explorer

```{r}
# ── CHANGE 9: Only show meaningful, renamed columns ───────────────────────────
# Previously dumped all raw columns, including cryptic coded names.
datatable(
  df_clean %>%
    select(
      Neighborhood  = name,
      `Poverty Rate (%)`    = poverty_rate,
      `% White`             = percent_white,
      `Obesity Rate (%)`    = obesity_rate,
      `Uninsured Rate (%)`  = uninsured_rate,
      `Obesity Count`       = hcsob_2023_2024,
      Population            = pop_2020_2024
    ) %>%
    mutate(across(where(is.numeric), ~round(.x, 2))),
  options  = list(pageLength = 8, scrollX = TRUE),
  rownames = FALSE
)
```
```

### Purpose
This interactive table ensures:
Transparency
Reproducibility
Direct engagement with raw data
Users can explore, filter, and validate all findings.

Row {data-height=300}
-----------------------------------------------------------------------

### Final Scientific Interpretation
Health disparities in Chicago emerge from systemic, interconnected factors:
Economic inequality drives differential exposure and access
Demographic patterns reflect historical structural inequities
Spatial clustering indicates place-based risk environments

### Conclusion
Chronic disease burden is not randomly distributed —
it is shaped by a combination of social, economic, and geographic determinants.
This analysis provides a framework for targeted, data-driven public health intervention
