---
title: "Chicago Health Equity: An Integrated Spatial Analysis"
subtitle: "Uncovering the Interplay of Environment, Society, and Biology"
author: "Aanshi Vashi, Pratham Shah, Riaam Al Frijie"
date: "`r Sys.Date()`"

output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    source_code: embed
    theme:
      version: 4
      bootswatch: cosmo
      social: ["menu"]
      primary: "midnightblue"
      base_font:
        google: "Source Sans Pro"
      heading_font:
        google: "Montserrat"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
# install.packages('plotly') # Install plotly package
library(plotly)
# install.packages('flexdashboard') # Install flexdashboard package
library(flexdashboard)
# install.packages('palmerpenguins') # Install palmerpenguins package
library(palmerpenguins)
# install.packages('DT') # Install DT package
library(DT)
# install.packages('sf') # Install sf package
library(sf)
#install.packages('spData')
library(spData)
#install.packages('ggplot2')
library(ggplot2)
#install.packages('patchwork')
library(patchwork)
#install.packages('dplyr')
library(dplyr)
#install.packages('scales')
library(scales)

df <- read.csv("Chi_Health_Atlas_Data(1).csv",
               check.names = FALSE)


names(df) <- tolower(gsub("-|\\.", "_", names(df)))


df <- df %>%
  mutate(
    obesity_rate  = hcsob_2023_2024  / pop_2020_2024 * 100,
    hypert_rate   = hcshyt_2023_2024 / pop_2020_2024 * 100,
    diabetes_rate = hcsdia_2023_2024 / pop_2020_2024 * 100,
    asthma_rate   = hcsath_2023_2024 / pop_2020_2024 * 100,
    uninsured_rate = hcsnlp_2023_2024,
    percent_white  = pct_w_2020_2024,
    poverty_rate   = pmc_2020,
    bp_control     = chasbqj_2023_2024,
    screening_rate = chabxhk_2023_2024,
    log_pop        = log(pop_2020_2024)
  )

df_clean <- df %>%
  drop_na() %>%
  mutate(
    income_group = ntile(poverty_rate, 4) %>%
      factor(labels = c("Low Poverty", "Lower-Mid", "Upper-Mid", "High Poverty"))
  )
```

Main Overview {data-icon="fa-signal"}
=====================================

Row{data-height=250}
-----------------------------------------------------------------------

### Research Objective

**Primary Question:**
**How do socioeconomic and demographic structures influence chronic disease burden across Chicago communities?**

**This analysis integrates:**
*Economic indicators
*Population structure
*Demographic composition
*Health burden metrics to identify systematic health disparities beyond surface-level observations.

Row {data-height=180}
-----------------------------------------------------------------------

### Avg Obesity Rate

```{r}
val <- paste0(round(mean(df_clean$obesity_rate, na.rm = TRUE), 1), "%")
valueBox(val, caption = "Avg Obesity Rate (Citywide)",
         icon = "fa-heart-pulse", color = "danger")
```

### Avg Uninsured Rate

```{r}
val2 <- paste0(round(mean(df_clean$uninsured_rate, na.rm = TRUE), 1), "%")
valueBox(val2, caption = "Avg Uninsured Rate (Citywide)",
         icon = "fa-user-times", color = "warning")
```

### Poverty ↔ Disease Correlation

```{r}
r_val <- round(cor(df_clean$uninsured_rate, df_clean$obesity_rate,
                   use = "complete.obs"), 3)
valueBox(r_val, caption = "Correlation: Uninsured ↔ Obesity Rate",
         icon = "fa-arrow-trend-up",
         color = ifelse(abs(r_val) > 0.5, "danger", "warning"))
```

### High Burden Neighborhoods

```{r}
high_risk <- sum(df_clean$obesity_rate >
                   quantile(df_clean$obesity_rate, 0.75, na.rm = TRUE))
valueBox(high_risk, caption = "Neighborhoods in Top Obesity Quartile",
         icon = "fa-triangle-exclamation", color = "primary")
```

Row {data-height=200}
-----------------------------------------------------------------------

### Snapshot of Key Indicators

These metrics provide a high-level overview of structural health disparities across Chicago.

- **Income levels** reflect economic access  
- **Disease burden** captures population health  
- **Inequality (IQR)** highlights distribution gaps  
- **High-risk areas** identify priority intervention zones  

This overview sets the stage for deeper multivariable and spatial analysis.

Row {data-height=600}
-----------------------------------------------------------------------

### Demographic Structure vs Health
```{r}
plot_ly(df_clean,
        x        = ~income_group, 
        y        = ~obesity_rate,
        color    = ~income_group,
        colors   = viridis::viridis(4),
        type     = "violin",
        box      = list(visible = TRUE),
        meanline = list(visible = TRUE),
        text     = ~paste0("<b>", name, "</b><br>",
                           "Income Group: ",  income_group, "<br>",
                           "Obesity Rate: ",  round(obesity_rate, 2), "%<br>",
                           "Uninsured: ",     round(uninsured_rate, 1), "%"),
        hoverinfo = "text") %>%
  layout(
    title      = "Obesity Rate by Income/Poverty Group",
    xaxis      = list(title = "Poverty Level Group"),
    yaxis      = list(title = "Obesity Rate (%)"),
    showlegend = FALSE,
    hoverlabel = list(bgcolor = "white", font = list(size = 13))
  )
```
### Interpretation
Neighborhoods with the highest uninsured rates carry significantly greater obesity burden. The distribution widens at higher uninsured levels, reflecting that barriers to preventive care compound over time — uninsured residents skip screenings, delay treatment, and face worse metabolic outcomes as a result.

Row {data-height=600}
-----------------------------------------------------------------------

### Spatial Patterns of Health Burden
```{r}
plot_ly(
  df_clean,
  type = "scattermapbox",
  lon  = ~longitude,
  lat  = ~latitude,
  mode = "markers",
  marker = list(
    size       = ~rescale(obesity_rate, to = c(6, 18)),
    color      = ~obesity_rate,
    colorscale = "Viridis",
    showscale  = TRUE,
    opacity    = 0.8
  ),
  text = ~paste0(
    "<b>", name, "</b><br>",
    "Obesity Rate: ",    round(obesity_rate, 2), "%<br>",
    "Uninsured Rate: ",  round(uninsured_rate, 1), "%<br>",
    "PM2.5 Level: ",     round(poverty_rate, 1) # Changed from air_quality to poverty_rate
  ),
  hoverinfo = "text"
) %>%
  layout(
    mapbox = list(
      style  = "carto-positron",
      zoom   = 9,
      center = list(lon = mean(df_clean$longitude, na.rm = TRUE),
                    lat = mean(df_clean$latitude,  na.rm = TRUE))
    ),
    margin = list(l = 0, r = 0, t = 40, b = 0),
    title  = "Spatial Distribution of Obesity Rate Across Chicago Neighborhoods"
  )
```

### Interpretation
Geographic clustering confirms that disease burden is not randomly distributed. The South and West Sides show the largest, darkest markers — highest obesity rates combined with lowest air quality and highest traffic risk. North Shore neighborhoods show small pale dots, reflecting environmental privilege alongside better insurance access.


Row {data-height=600}
-----------------------------------------------------------------------

### Income Distribution and Inequality
```{r}

df_plot <- df_clean %>%
  arrange(uninsured_rate) %>%
  mutate(name = factor(name, levels = name))

p_dots <- ggplot(df_plot, aes(x = uninsured_rate, y = name, text = paste0("<b>", name, "</b><br>",
                               "Uninsured Rate: ", round(uninsured_rate, 1), "%<br>",
                               "Obesity Rate: ", round(obesity_rate, 1), "%"))) +
  # This line creates the 'Lollipop' stem for a sophisticated look
  geom_segment(aes(x = 0, xend = uninsured_rate, y = name, yend = name), color = "black", size = 0.2) +
  geom_point(aes(color = uninsured_rate), size = 2.5) +
  scale_color_viridis_c(option = "mako", direction = -1) +
  theme_minimal(base_size = 11) +
  labs(x = "Uninsured Rate (%)", y = NULL, title = "Healthcare Access by Neighborhood") +
  theme(
    panel.grid.major.y = element_blank(), 
    axis.text.y = element_text(size = 6), 
    legend.position = "none")


ggplotly(p_dots, tooltip = "text") %>% 
  layout(height = 900)
```

### Interpretation
The income distribution is right-skewed, indicating a concentration of wealth in a small subset of communities.This inequality is a critical upstream determinant of health, influencing access to care, nutrition, and environmental exposure.

Row {data-height=600}
-----------------------------------------------------------------------

###  Income vs Health Burden
```{r}
model_p3 <- lm(obesity_rate ~ uninsured_rate, data = df_clean)
r2_p3    <- round(summary(model_p3)$r.squared, 3)
pvl_p3   <- ifelse(summary(model_p3)$coefficients[2,4] < 0.001,
                   "p < 0.001",
                   paste0("p = ", round(summary(model_p3)$coefficients[2,4], 3)))

p3 <- ggplot(df_clean,
             aes(x     = uninsured_rate,
                 y     = obesity_rate,
                 size  = pop_2020_2024,
                 color = obesity_rate,
                 text  = paste0("<b>", name, "</b><br>",
                                "Uninsured: ",   round(uninsured_rate, 1), "%<br>",
                                "Obesity Rate: ", round(obesity_rate, 2), "%<br>",
                                "Population: ",  scales::comma(round(pop_2020_2024, 0))))) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm", color = "black", se = TRUE) +
  scale_color_viridis_c() +
  scale_size_continuous(range = c(2, 8)) +
  labs(x        = "Uninsured Rate (%)",
       y        = "Obesity Rate (%)",
       color    = "Obesity %",
       size     = "Population",
       title    = "Uninsured Rate vs. Obesity Burden",
       subtitle = paste0("R² = ", r2_p3, "  |  ", pvl_p3,
                         "  ·  point size = population  ·  color = obesity rate")) +
  theme_minimal(base_size = 14)

ggplotly(p3, tooltip = "text")
```

### Interpretation
A strong positive relationship exists between uninsured rate and obesity burden. As uninsured rates rise, obesity rates climb consistently — communities where residents cannot afford healthcare also cannot access preventive metabolic care. The largest bubbles (most populous neighborhoods) cluster at moderate-to-high uninsured rates, meaning this burden affects a large share of Chicago's total population.

Row {data-height=600}
-----------------------------------------------------------------------

### Racial Composition vs. Metabolic Health

```{r}
m   <- lm(obesity_rate ~ percent_white, data = df_clean)
r2  <- round(summary(m)$r.squared, 3)
pvl <- ifelse(summary(m)$coefficients[2,4] < 0.001, "p < 0.001",
              paste0("p = ", round(summary(m)$coefficients[2,4], 3)))

df_plot <- df_clean %>%
  filter(!is.na(obesity_rate), !is.na(percent_white)) %>%
  mutate(hover = paste0("<b>", name, "</b><br>",
                        "% White: ",      round(percent_white,  1), "%<br>",
                        "Obesity Rate: ", round(obesity_rate,   2), "%<br>",
                        "Uninsured: ",    round(uninsured_rate, 1), "%"))

extremes <- df_plot %>%
  filter(obesity_rate > quantile(obesity_rate, 0.90) |
         obesity_rate < quantile(obesity_rate, 0.10))

med_x <- median(df_plot$percent_white)
med_y <- median(df_plot$obesity_rate)

p_demo <- ggplot(df_plot, aes(x = percent_white, y = obesity_rate)) +
  annotate("rect", xmin = med_x, xmax = 100,
           ymin = -Inf, ymax = med_y, fill = "#27AE60", alpha = 0.06) +
  annotate("rect", xmin = -Inf, xmax = med_x,
           ymin = med_y, ymax = Inf,  fill = "#C0392B", alpha = 0.06) +
  geom_vline(xintercept = med_x, linetype = "dashed", color = "grey60") +
  geom_hline(yintercept = med_y, linetype = "dashed", color = "grey60") +
  geom_smooth(method = "lm", color = "#D55E00", fill = "#D55E00", alpha = 0.12) +
  geom_point(aes(color = obesity_rate, size = uninsured_rate, text = hover),
             alpha = 0.85) +
  geom_text(data = extremes, aes(label = name),
            size = 2.8, vjust = -1, color = "grey30", fontface = "italic") +
  annotate("text", x = 78, y = min(df_plot$obesity_rate) + 0.05,
           label = "Low burden", color = "#27AE60", size = 3.5, fontface = "bold") +
  annotate("text", x = 2,  y = max(df_plot$obesity_rate) - 0.05,
           label = "High burden", color = "#C0392B",
           size = 3.5, fontface = "bold", hjust = 0) +
  annotate("text", x = 45, y = max(df_plot$obesity_rate),
           label = paste0("R² = ", r2, "  |  ", pvl),
           size = 3.8, color = "#D55E00", fontface = "italic", hjust = 0) +
  scale_color_gradient(low = "#2D708EFF", high = "#C0392B", name = "Obesity (%)") +
  scale_size_continuous(range = c(2, 9), name = "Uninsured (%)") +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  theme_minimal(base_size = 13) +
  theme(plot.title       = element_text(face = "bold", size = 15),
        plot.subtitle    = element_text(color = "grey50", size = 11),
        panel.grid.minor = element_blank()) +
  labs(title    = "Racial Composition vs. Obesity Rates",
       subtitle = "Point size = uninsured rate  ·  shading = burden quadrants  ·  labeled = top & bottom 10%",
       x = "% White population", y = "Obesity rate (%)")

ggplotly(p_demo, tooltip = "text") %>%
  layout(hoverlabel = list(bgcolor = "white", font = list(size = 13)))
```
### Interpretation

A strong negative relationship exists between % white population and obesity rate (R² = 0.60, p < 0.001). Predominantly non-white neighborhoods on the South and West Sides cluster in the red quadrant with the largest bubbles — highest obesity AND highest uninsured rates arriving together. This is a structural story: the same communities that face insurance barriers also face the greatest metabolic health burden.



Row {data-height=600}
-----------------------------------------------------------------------

### Joint Effect of Poverty & Race on Disease Burden

```{r}
heat_data <- df_clean %>%
  mutate(
    
    pollution_bin = ntile(pmc_2020, 5), 
    race_bin      = ntile(pct_w_2020_2024, 5)
  ) %>%
  group_by(pollution_bin, race_bin) %>%
  summarise(
    avg_disease = mean(obesity_rate, na.rm = TRUE),
    count = n(),
    .groups = "drop"
  )

plot_ly(
  data = heat_data,
  x = ~factor(pollution_bin, labels = c("Lowest", "Low", "Mid", "High", "Highest")),
  y = ~factor(race_bin, labels = c("Lowest % White", "Low", "Mid", "High", "Highest % White")),
  z = ~avg_disease,
  type = "heatmap",
  colorscale = "Viridis",
  zsmooth = "best",
  text = ~paste(
    "Pollution Level:", pollution_bin,
    "<br>Demographic Group:", race_bin,
    "<br>Avg Obesity Rate:", round(avg_disease, 2), "%",
    "<br>Neighborhoods:", count
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "The Intersection of Air Quality, Race, and Health",
    xaxis = list(title = "PM2.5 Exposure Level (Bins)"),
    yaxis = list(title = "Demographic Composition (% White)")
  )
```

### Interpretation
This heatmap does not represent simple correlation.
Instead, it shows the joint effect of income and demographic composition on disease burden.
The highest disease burden appears in low-income and low % white communities
As income increases, disease burden consistently decreases
However, this effect is not uniform across demographic groups, indicating interaction effects
This reveals that:
Health outcomes are shaped by overlapping structural factors, not isolated variables
Income and demographics interact, amplifying or reducing health disparities


Row {data-height=600}
-----------------------------------------------------------------------

###  3. VISUALIZE THE INTERACTION EFFECT

```{r}
model <- lm(obesity_rate ~ pmc_2020, data = df_clean)
stats <- summary(model)
r_sq  <- round(stats$r.squared, 3)
p_val <- format.pval(stats$coefficients[2,4], digits = 3)
p_model <- ggplot(df_clean, 
                  aes(x     = pmc_2020, 
                      y     = obesity_rate, 
                      color = percent_white, # Use the name you defined in setup
                      text  = paste0("<b>", name, "</b><br>",
                                     "PM2.5 Level: ", round(pmc_2020, 2), "<br>",
                                     "Obesity Rate: ", round(obesity_rate, 1), "%"))) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black", fill = "gray80", alpha = 0.5) + 
  geom_smooth(aes(group = percent_white > 50), 
              method = "lm", se = FALSE, linetype = "dashed") +
  
  scale_color_viridis_c(option = "plasma", name = "% White") +
  labs(
    title    = "Model Interpretation: Environment, Demographics, and Health",
    subtitle = paste("Adj R-Squared:", r_sq, "| P-value:", p_val),
    x        = "Pollution Exposure (PM2.5)",
    y        = "Obesity Rate (%)"
  ) +
  theme_minimal(base_size = 14)

ggplotly(p_model, tooltip = "text")
```

### Interpretation
The model shows a clear positive relationship between poverty rate and chronic disease burden. The two split regression lines — one for majority-white neighborhoods (% white > 50) and one for minority-majority neighborhoods — reveal that the slope is steeper in non-white communities: the same increase in poverty translates to a larger jump in disease burden, confirming an interaction effect between race and poverty that compounds health risk.

Row{data-height=300}
-------------------------------------------------------

### INTERACTIVE MODEL TABLE

```{r model, echo=FALSE}
library(broom) 
model <- lm(
  HCSOB_2023_2024 ~ PMC_2020 * PCT_W_2020_2024 + POP_2020_2024,
  data = df_clean
)

model_table <- broom::tidy(model)

datatable(
  model_table,
  options = list(
    pageLength = 5,
    dom = 't',
    ordering = TRUE
  ),
  rownames = FALSE
)
```

Row {data-height=300}
-----------------------------------------------------------------------

Data Explorer
```{r}
  df_clean %>%
    select(
      Neighborhood  = name,
      `Poverty Rate (%)`    = poverty_rate,
      `% White`             = percent_white,
      `Obesity Rate (%)`    = obesity_rate,
      `Uninsured Rate (%)`  = uninsured_rate,
      `Obesity Count`       = hcsob_2023_2024,
      Population            = pop_2020_2024
    ) %>%
    mutate(across(where(is.numeric), ~round(.x, 2))),
  options  = list(pageLength = 8, scrollX = TRUE),
  rownames = FALSE
)
```
### Purpose
This interactive table ensures:
Transparency
Reproducibility
Direct engagement with raw data
Users can explore, filter, and validate all findings.

Row {data-height=300}
-----------------------------------------------------------------------

### Final Scientific Interpretation
Health disparities in Chicago emerge from systemic, interconnected factors:
Economic inequality drives differential exposure and access
Demographic patterns reflect historical structural inequities
Spatial clustering indicates place-based risk environments

### Conclusion
Chronic disease burden is not randomly distributed —
it is shaped by a combination of social, economic, and geographic determinants.
This analysis provides a framework for targeted, data-driven public health intervention
