---
title: "Chicago Health Equity: An Integrated Spatial Analysis"
subtitle: "Uncovering the Interplay of Environment, Society, and Biology"
author: "Aanshi Vashi, Pratham Shah, Riaam Al Frijie"
date: "`r Sys.Date()`"

output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    source_code: embed
    theme:
      version: 4
      bootswatch: cosmo
      social: ["menu"]
      primary: "midnightblue"
      base_font:
        google: "Source Sans Pro"
      heading_font:
        google: "Montserrat"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
# install.packages('plotly') # Install plotly package
library(plotly)
# install.packages('flexdashboard') # Install flexdashboard package
library(flexdashboard)
# install.packages('palmerpenguins') # Install palmerpenguins package
library(palmerpenguins)
# install.packages('DT') # Install DT package
library(DT)
# install.packages('sf') # Install sf package
library(sf)
#install.packages('spData')
library(spData)
#install.packages('ggplot2')
library(ggplot2)
#install.packages('patchwork')
library(patchwork)
#install.packages('dplyr')
library(dplyr)
#install.packages('scales')
library(scales)

# LOAD DATA
df <- read.csv("Chi_Health_Atlas_Data.csv")

# CLEAN COLUMN NAMES
names(df) <- gsub("-", "_", names(df))
names(df) <- gsub("\\.", "_", names(df))

# DEBUG: SEE COLUMN NAMES
print(colnames(df))

# CLEAN + TRANSFORM
df_clean <- df %>%
  drop_na() %>%
  mutate(
    log_income = log(PMC_2020 + 1),
    log_pop = log(POP_2020_2024)
  )
```

Main Overview {data-icon="fa-signal"}
=====================================

Row{data-height=250}
-----------------------------------------------------------------------

### Research Objective

**Primary Question:**
**How do socioeconomic and demographic structures influence chronic disease burden across Chicago communities?**

**This analysis integrates:**
*Economic indicators
*Population structure
*Demographic composition
*Health burden metrics to identify systematic health disparities beyond surface-level observations.

Row {data-height=180}
-----------------------------------------------------------------------
### Value box 1

```{r}
val_income <- scales::dollar(median(df_clean$PMC_2020, na.rm = TRUE))

valueBox(
  value = val_income,
  caption = "Median Income (Citywide)",
  icon = "fa-money-bill-wave",
  color = "primary"
)
```

### Value box 2

```{r}
avg_disease <- round(mean(df_clean$HCSOB_2023_2024, na.rm = TRUE), 1)

valueBox(
  value = avg_disease,
  caption = "Average Chronic Disease Burden",
  icon = "fa-heart-pulse",
  color = "danger"
)
```

### Value box 3

```{r}
iqr_income <- scales::dollar(IQR(df_clean$PMC_2020, na.rm = TRUE))

valueBox(
  value = iqr_income,
  caption = "Income Inequality (IQR)",
  icon = "fa-chart-line",
  color = "warning"
)
```

### Value box 4

```{r}
high_risk <- sum(df_clean$HCSOB_2023_2024 > quantile(df_clean$HCSOB_2023_2024, 0.75, na.rm = TRUE))

valueBox(
  value = high_risk,
  caption = "High Disease Burden Areas",
  icon = "fa-triangle-exclamation",
  color = "success"
)
```

Row {data-height=200}
-----------------------------------------------------------------------

### 🔍 Snapshot of Key Indicators

These metrics provide a high-level overview of structural health disparities across Chicago.

- **Income levels** reflect economic access  
- **Disease burden** captures population health  
- **Inequality (IQR)** highlights distribution gaps  
- **High-risk areas** identify priority intervention zones  

This overview sets the stage for deeper multivariable and spatial analysis.

Row {data-height=600}
-----------------------------------------------------------------------

### Demographic Structure vs Health
```{r}
df_clean <- df_clean %>%
  mutate(
    income_group = ntile(PMC_2020, 4) %>%
      factor(labels = c("Low", "Lower-Mid", "Upper-Mid", "High"))
  )

p4 <- ggplot(df_clean, aes(x = income_group, y = HCSOB_2023_2024, fill = income_group)) +
  geom_violin(alpha = 0.6) +
  geom_boxplot(width = 0.2, fill = "white") +
  scale_fill_viridis_d() +
  labs(
    x = "Income Group",
    y = "Chronic Disease Burden",
    title = "Disease Burden Across Income Groups"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

ggplotly(p4)
```
### Interpretation
Demographic composition shows strong alignment with health disparities.
This reflects structural and historical inequalities rather than inherent biological differences.

Row {data-height=600}
-----------------------------------------------------------------------

### Spatial Patterns of Health Burden
```{r}
sf_data <- st_as_sf(df_clean, coords = c("Longitude", "Latitude"), crs = 4326)

plot_ly(
  df_clean,
  type = "scattermapbox",
  lon = ~Longitude,
  lat = ~Latitude,
  mode = "markers",
  marker = list(
    size = ~scales::rescale(POP_2020_2024, to = c(6, 18)),
    color = ~HCSOB_2023_2024,
    colorscale = "Viridis",
    showscale = TRUE,
    opacity = 0.8
  ),
  text = ~paste0(
    "<b>", Name, "</b><br>",
    "Disease Burden: ", round(HCSOB_2023_2024, 2), "<br>",
    "Income: ", scales::dollar(PMC_2020), "<br>",
    "Population: ", scales::comma(POP_2020_2024)
  ),
  hoverinfo = "text"
) %>%
  layout(
    mapbox = list(
      style = "carto-positron",
      zoom = 9,
      center = list(
        lon = mean(df_clean$Longitude),
        lat = mean(df_clean$Latitude)
      )
    ),
    margin = list(l = 0, r = 0, t = 40, b = 0),
    title = "Interactive Spatial Distribution of Chronic Disease Burden"
  )
```

### Interpretation
Geographic clustering reveals that disease burden is spatially structured.
This suggests localized drivers such as environmental exposure, healthcare accessibility, and neighborhood-level inequality.


Row {data-height=600}
-----------------------------------------------------------------------

### Income Distribution and Inequality
```{r}
p1 <- ggplot(df_clean, aes(x = "All Communities", y = PMC_2020)) +
  geom_violin(fill = "#2C7FB8", alpha = 0.6, color = NA) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(width = 0.08, alpha = 0.3, color = "#08306B") +
  labs(
    y = "Median Per Capita Income",
    x = NULL,
    title = "Income Distribution Across Communities"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_blank())

ggplotly(p1)
```
### Interpretation
The income distribution is right-skewed, indicating a concentration of wealth in a small subset of communities.This inequality is a critical upstream determinant of health, influencing access to care, nutrition, and environmental exposure.

Row {data-height=600}
-----------------------------------------------------------------------

###  Income vs Health Burden
```{r}
p3 <- ggplot(df_clean, 
             aes(x = PMC_2020, 
                 y = HCSOB_2023_2024,
                 size = POP_2020_2024,
                 color = HCSOB_2023_2024,
                 text = paste("Community:", Name))) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm", color = "black", se = TRUE) +
  scale_color_viridis_c() +
  scale_size_continuous(range = c(2, 8)) +
  labs(
    x = "Income",
    y = "Chronic Disease Burden",
    color = "Disease",
    size = "Population",
    title = "Income vs Disease Burden with Population Context"
  ) +
  theme_minimal(base_size = 14)

ggplotly(p3, tooltip = "text")
```

### Interpretation
A clear inverse relationship exists between income and chronic disease burden.
Lower-income communities experience higher disease prevalence, supporting well-established epidemiological patterns linking socioeconomic stressors to biological outcomes.


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
-----------------------------------------------------------------------


### Pollution and Obesity Relationship
```{r}
# Scatterplot: Pollution vs Obesity

# Value box 1
obesity_label <- paste0(round(mean(df$HCSOB_2023_2024, na.rm = TRUE), 1), "%")
valueBox(obesity_label, icon = "fa-weight-scale", color = "warning")

# Value box 2
high_pollution_count <- sum(df$PMC_2020 > 9.5, na.rm = TRUE)
valueBox(high_pollution_count, icon = "fa-smog", color = "danger")

# Value box 3
valueBox(nrow(df), icon = "fa-building-columns", color = "primary")

# plot 1
p1 <- ggplot(df, aes(x = PMC_2020, y = HCSOB_2023_2024)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal()

ggplotly(p1)

# Linear regression model 1
model <- lm(HCSOB_2023_2024 ~ PMC_2020, data = df)

# Show results
summary(model)

# plot 2
p2 <- ggplot(df, aes(x = PCT_W_2020_2024, y = HCSOB_2023_2024)) +
  geom_point(color = "darkgreen", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "black") +
  theme_minimal()

ggplotly(p2)

# Linear regression model: Demographics ~ Obesity
model2 <- lm(HCSOB_2023_2024 ~ PMC_2020 + PCT_W_2020_2024, data = df)

# Show results
summary(model2)
```

### Interpretation
Air pollution becomes more meaningful when combined with demographic factors. This suggests health outcomes are shaped by both environmental and social conditions rather than a single variable.
