---
title: "Chicago Health Equity: An Integrated Spatial Analysis"
subtitle: "Uncovering the Interplay of Environment, Society, and Biology"
author: "Aanshi Vashi, Pratham Shah, Riaam Al Frijie"
date: "`r Sys.Date()`"

output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    source_code: embed
    theme:
      version: 4
      bootswatch: cosmo
      social: ["menu"]
      primary: "midnightblue"
      base_font:
        google: "Source Sans Pro"
      heading_font:
        google: "Montserrat"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
# install.packages('plotly') # Install plotly package
library(plotly)
# install.packages('flexdashboard') # Install flexdashboard package
library(flexdashboard)
# install.packages('palmerpenguins') # Install palmerpenguins package
library(palmerpenguins)
# install.packages('DT') # Install DT package
library(DT)
# install.packages('sf') # Install sf package
library(sf)
#install.packages('spData')
library(spData)
#install.packages('ggplot2')
library(ggplot2)
#install.packages('patchwork')
library(patchwork)
#install.packages('dplyr')
library(dplyr)
#install.packages('scales')
library(scales)

# LOAD DATA
df <- read.csv("Chi_Health_Atlas_Data(1).csv")

# CLEAN COLUMN NAMES
names(df) <- gsub("-", "_", names(df))
names(df) <- gsub("\\.", "_", names(df))

# DEBUG: SEE COLUMN NAMES
print(colnames(df))

# CLEAN + TRANSFORM
df_clean <- df %>%
  drop_na() %>%
  mutate(
    log_income = log(PMC_2020 + 1),
    log_pop = log(POP_2020_2024)
  )
```

Main Overview {data-icon="fa-signal"}
=====================================

Row{data-height=250}
-----------------------------------------------------------------------

### Research Objective

**Primary Question:**
**How do socioeconomic and demographic structures influence chronic disease burden across Chicago communities?**

**This analysis integrates:**
*Economic indicators
*Population structure
*Demographic composition
*Health burden metrics to identify systematic health disparities beyond surface-level observations.

Row {data-height=180}
-----------------------------------------------------------------------
### Value box 1

```{r}
val_income <- scales::dollar(median(df_clean$PMC_2020, na.rm = TRUE))

valueBox(
  value = val_income,
  caption = "Median Income (Citywide)",
  icon = "fa-money-bill-wave",
  color = "primary"
)
```

### Value box 2

```{r}
avg_disease <- round(mean(df_clean$HCSOB_2023_2024, na.rm = TRUE), 1)

valueBox(
  value = avg_disease,
  caption = "Average Chronic Disease Burden",
  icon = "fa-heart-pulse",
  color = "danger"
)
```

### Value box 3

```{r}
iqr_income <- scales::dollar(IQR(df_clean$PMC_2020, na.rm = TRUE))

valueBox(
  value = iqr_income,
  caption = "Income Inequality (IQR)",
  icon = "fa-chart-line",
  color = "warning"
)
```

### Value box 4

```{r}
high_risk <- sum(df_clean$HCSOB_2023_2024 > quantile(df_clean$HCSOB_2023_2024, 0.75, na.rm = TRUE))

valueBox(
  value = high_risk,
  caption = "High Disease Burden Areas",
  icon = "fa-triangle-exclamation",
  color = "success"
)
```

Row {data-height=200}
-----------------------------------------------------------------------

### 🔍 Snapshot of Key Indicators

These metrics provide a high-level overview of structural health disparities across Chicago.

- **Income levels** reflect economic access  
- **Disease burden** captures population health  
- **Inequality (IQR)** highlights distribution gaps  
- **High-risk areas** identify priority intervention zones  

This overview sets the stage for deeper multivariable and spatial analysis.

Row {data-height=600}
-----------------------------------------------------------------------

### Demographic Structure vs Health
```{r}
df_clean <- df_clean %>%
  mutate(
    income_group = ntile(PMC_2020, 4) %>%
      factor(labels = c("Low", "Lower-Mid", "Upper-Mid", "High"))
  )

p4 <- ggplot(df_clean, aes(x = income_group, y = HCSOB_2023_2024, fill = income_group)) +
  geom_violin(alpha = 0.6) +
  geom_boxplot(width = 0.2, fill = "white") +
  scale_fill_viridis_d() +
  labs(
    x = "Income Group",
    y = "Chronic Disease Burden",
    title = "Disease Burden Across Income Groups"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

ggplotly(p4)
```
### Interpretation
Demographic composition shows strong alignment with health disparities.
This reflects structural and historical inequalities rather than inherent biological differences.

Row {data-height=600}
-----------------------------------------------------------------------

### Spatial Patterns of Health Burden
```{r}
sf_data <- st_as_sf(df_clean, coords = c("Longitude", "Latitude"), crs = 4326)

plot_ly(
  df_clean,
  type = "scattermapbox",
  lon = ~Longitude,
  lat = ~Latitude,
  mode = "markers",
  marker = list(
    size = ~scales::rescale(POP_2020_2024, to = c(6, 18)),
    color = ~HCSOB_2023_2024,
    colorscale = "Viridis",
    showscale = TRUE,
    opacity = 0.8
  ),
  text = ~paste0(
    "<b>", Name, "</b><br>",
    "Disease Burden: ", round(HCSOB_2023_2024, 2), "<br>",
    "Income: ", scales::dollar(PMC_2020), "<br>",
    "Population: ", scales::comma(POP_2020_2024)
  ),
  hoverinfo = "text"
) %>%
  layout(
    mapbox = list(
      style = "carto-positron",
      zoom = 9,
      center = list(
        lon = mean(df_clean$Longitude),
        lat = mean(df_clean$Latitude)
      )
    ),
    margin = list(l = 0, r = 0, t = 40, b = 0),
    title = "Interactive Spatial Distribution of Chronic Disease Burden"
  )
```

### Interpretation
Geographic clustering reveals that disease burden is spatially structured.
This suggests localized drivers such as environmental exposure, healthcare accessibility, and neighborhood-level inequality.


Row {data-height=600}
-----------------------------------------------------------------------

### Income Distribution and Inequality
```{r}
p1 <- ggplot(df_clean, aes(x = "All Communities", y = PMC_2020)) +
  geom_violin(fill = "#2C7FB8", alpha = 0.6, color = NA) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(width = 0.08, alpha = 0.3, color = "#08306B") +
  labs(
    y = "Median Per Capita Income",
    x = NULL,
    title = "Income Distribution Across Communities"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_blank())

ggplotly(p1)
```
### Interpretation
The income distribution is right-skewed, indicating a concentration of wealth in a small subset of communities.This inequality is a critical upstream determinant of health, influencing access to care, nutrition, and environmental exposure.

Row {data-height=600}
-----------------------------------------------------------------------

###  Income vs Health Burden
```{r}
p3 <- ggplot(df_clean, 
             aes(x = PMC_2020, 
                 y = HCSOB_2023_2024,
                 size = POP_2020_2024,
                 color = HCSOB_2023_2024,
                 text = paste("Community:", Name))) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm", color = "black", se = TRUE) +
  scale_color_viridis_c() +
  scale_size_continuous(range = c(2, 8)) +
  labs(
    x = "Income",
    y = "Chronic Disease Burden",
    color = "Disease",
    size = "Population",
    title = "Income vs Disease Burden with Population Context"
  ) +
  theme_minimal(base_size = 14)

ggplotly(p3, tooltip = "text")
```

### Interpretation
A clear inverse relationship exists between income and chronic disease burden.
Lower-income communities experience higher disease prevalence, supporting well-established epidemiological patterns linking socioeconomic stressors to biological outcomes.


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
-----------------------------------------------------------------------


### Pollution and Obesity Relationship
```{r}
# Scatterplot: Pollution vs Obesity

# Value box 1
obesity_label <- paste0(round(mean(df$HCSOB_2023_2024, na.rm = TRUE), 1), "%")
valueBox(obesity_label, icon = "fa-weight-scale", color = "warning")

# Value box 2
high_pollution_count <- sum(df$PMC_2020 > 9.5, na.rm = TRUE)
valueBox(high_pollution_count, icon = "fa-smog", color = "danger")

# Value box 3
valueBox(nrow(df), icon = "fa-building-columns", color = "primary")

# plot 1
p1 <- ggplot(df, aes(x = PMC_2020, y = HCSOB_2023_2024)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal()

ggplotly(p1)

# Linear regression model 1
model <- lm(HCSOB_2023_2024 ~ PMC_2020, data = df)

# Show results
summary(model)

# plot 2
p2 <- ggplot(df, aes(x = PCT_W_2020_2024, y = HCSOB_2023_2024)) +
  geom_point(color = "darkgreen", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "black") +
  theme_minimal()

ggplotly(p2)

# Linear regression model: Demographics ~ Obesity
model2 <- lm(HCSOB_2023_2024 ~ PMC_2020 + PCT_W_2020_2024, data = df)

# Show results
summary(model2)
```

### Interpretation
Air pollution becomes more meaningful when combined with demographic factors. This suggests health outcomes are shaped by both environmental and social conditions rather than a single variable.

Data Explorer {data-icon="fa-database"}
=====================================

Row {data-height=600}
-----------------------------------------------------------------------

### Correlation Structure

```{r}
heat_data <- df_clean %>%
  mutate(
    income_bin = ntile(PMC_2020, 5),
    race_bin = ntile(PCT_W_2020_2024, 5)
  ) %>%
  group_by(income_bin, race_bin) %>%
  summarise(
    avg_disease = mean(HCSOB_2023_2024),
    count = n(),
    .groups = "drop"
  )

plot_ly(zsmooth = "best",
  heat_data,
  x = ~factor(income_bin, labels = c("Lowest", "Low", "Mid", "High", "Highest")),
  y = ~factor(race_bin, labels = c("Lowest % White", "Low", "Mid", "High", "Highest % White")),
  z = ~avg_disease,
  type = "heatmap",
  colorscale = "Viridis",
  
  text = ~paste(
    "Income Group:", income_bin,
    "<br>Race Group:", race_bin,
    "<br>Avg Disease:", round(avg_disease, 2),
    "<br>Communities:", count
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Overlapping Effects of Income and Demographics on Disease Burden",
    xaxis = list(title = "Income Level"),
    yaxis = list(title = "Demographic Composition (% White)")
  )
```

### Interpretation
This heatmap does not represent simple correlation.
Instead, it shows the joint effect of income and demographic composition on disease burden.
Key insights:
The highest disease burden appears in low-income and low % white communities
As income increases, disease burden consistently decreases
However, this effect is not uniform across demographic groups, indicating interaction effects
This reveals that:
Health outcomes are shaped by overlapping structural factors, not isolated variables
Income and demographics interact, amplifying or reducing health disparities


Row {data-height=600}
-----------------------------------------------------------------------

###  3. VISUALIZE THE INTERACTION EFFECT

```{r}

model <- lm(HCSOB_2023_2024 ~ PMC_2020, data = df_clean)
stats <- summary(model)


r_sq <- round(stats$r.squared, 3)
p_val <- format.pval(stats$coefficients[2,4], digits = 3)

p_model <- ggplot(df_clean, 
                  aes(x = PMC_2020, 
                      y = HCSOB_2023_2024, 
                      color = PCT_W_2020_2024,
                      text = paste("Name:", Name, "<br>Income:", PMC_2020, "<br>Burden:", HCSOB_2023_2024))) +
  geom_point(size = 3, alpha = 0.7) +
  # se = TRUE adds the 'significance' ribbon (Confidence Interval)
  geom_smooth(method = "lm", se = TRUE, color = "black", fill = "gray80", alpha = 0.5) + 
  scale_color_viridis_c(option = "plasma") +
geom_smooth(aes(group = PCT_W_2020_2024 > 50), method = "lm", se = FALSE)+
  labs(
    title = "Model Interpretation: Income, Demographics, and Disease Burden",
    subtitle = paste("Adj R-Squared:", r_sq, "| P-value:", p_val),
    x = "Income",
    y = "Chronic Disease Burden",
    color = "% White"
  ) +
  theme_minimal(base_size = 14)

ggplotly(p_model, tooltip = "text")
```
### Interpretation
The model demonstrates a clear inverse relationship between income and chronic disease burden, as indicated by the negative slope of the regression line. Communities with higher median income generally exhibit lower levels of chronic disease, supporting well-established links between socioeconomic status and health outcomes.
The reported R-squared value suggests that income explains a meaningful portion of the variability in disease burden

Row{data-height=300}
-------------------------------------------------------

### INTERACTIVE MODEL TABLE

```{r model, echo=FALSE}
library(broom)

model <- lm(
  HCSOB_2023_2024 ~ PMC_2020 * PCT_W_2020_2024 + POP_2020_2024,
  data = df_clean
)

model_table <- broom::tidy(model)

datatable(
  model_table,
  options = list(
    pageLength = 5,
    dom = 't',
    ordering = TRUE
  ),
  rownames = FALSE
)
```
Row {data-height=300}
-----------------------------------------------------------------------

Data Explorer
```{r}
datatable(df_clean,
          options = list(pageLength = 8, scrollX = TRUE),
          rownames = FALSE)
```

### Purpose
This interactive table ensures:
Transparency
Reproducibility
Direct engagement with raw data
Users can explore, filter, and validate all findings.

Row {data-height=300}
-----------------------------------------------------------------------

### Final Scientific Interpretation
Health disparities in Chicago emerge from systemic, interconnected factors:
Economic inequality drives differential exposure and access
Demographic patterns reflect historical structural inequities
Spatial clustering indicates place-based risk environments

### Conclusion
Chronic disease burden is not randomly distributed —
it is shaped by a combination of social, economic, and geographic determinants.
This analysis provides a framework for targeted, data-driven public health intervention
